---
title: "Final project"
author: "Xinyun (Stacy) Li, Qiuyue Kong, Wenxu Liu, Siwei Stella Zhu"
date: "12/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview and Motivation

As future leaders in the field of public health, we are empowered by the resources available to us from HSPH. We hope to use our analytic skills to contribute to timely public health issues. Given that our daily life in the past two years has been severely impacted by the COVID-19 pandemic, we hope to produce new findings on how scientific research and public notice have impacted vaccine preference and vaccination rates’ impacts on cases and deaths.

# Related Work

# Initial Questions

1. How does vaccination distribution and administration change as delta variant becomes the dominant variant in the United States? 
2. How does public preference of different COVID-19 vaccination types change by new research outcomes or public notice? 
3. What proportion of people have been fully vaccinated? What proportion of people have only got one dose of the vaccine? What proportion of people have the booster? Are there any differences between age group (12+, 18+ and 65+ population)? 
4. How does the vaccination status (fully vaccinated, vaccinated with the first dose, vaccinated with booster dose) predict the number of cases and deaths in the future?  
5. How does different vaccination types affect cases & deaths, for the total population in the U.S. and populations among different states? 

# Data

### Sources

We used two COVID-related datasets collected from CDC website (open-source data download). We combined the datasets based on date and state and filter out meaningful variables for the purpose of this research. 

1. COVID-19 Vaccine Delivered and Administration Data - “Covid-19 vaccination in the United States, Jurisdiction”. [Link](https://data.cdc.gov/Vaccinations/COVID-19-Vaccinations-in-the-United-States-Jurisdi/unsk-b7fc )

2. “United States COVID-19 Cases and Deaths by State over Time.” [Link](https://data.cdc.gov/Case-Surveillance/United-States-COVID-19-Cases-and-Deaths-by-State-o/9mfq-cb36 )

### Data Cleanup

```{r}
library(tidyverse)
library(dplyr)
library(lubridate)
library(caret)
library(MASS)
library(rpart)
library(randomForest)
library(zoo)
library(maps)
library(gridExtra)
```

```{r}
# Import Data
vaccine<-read.csv("COVID-19_Vaccinations_in_the_United_States_Jurisdiction.csv")
case_death<-read.csv("United_States_COVID-19_Cases_and_Deaths_by_State_over_Time.csv")
```

```{r}
# Clean the case-death data into 50 states
USstate=c("AL","AK","AZ","AR","CA","CO","CT","DE","FL","GA","HI","ID","IL","IN","IA","KS",
          "KY","LA","ME","MD","MA","MI","MN","MS","MO","MT","NE","NV","NH","NJ","NM","NY",
          "NC","ND","OH","OK","OR","PA","RI","SC","SD","TN","TX","UT","VT","VA","WA","WV",
          "WI","WY")
case_death<-case_death %>% 
  transmute(date=mdy(submission_date),
            state=state,
            tot_case=tot_cases,
            tot_death=tot_death,
            new_case=new_case,
            new_death=new_death
            ) %>%
  filter(state %in% USstate) %>%
  filter(date >= as.Date("2021-01-01") & date<=as.Date("2021-11-30"))
```

```{r}
# Clean the vaccine data (change date to mdy)
vaccine<-vaccine %>% 
  dplyr::select(c(1,3,13,14,15,16,17,18,19,26,28,30,32,34,36,38,40,42,43,44,46,47,48,50,51,52,54,55,56,58,60,62,64,66,67,68)) %>%
  rename(state=Location) %>% 
  mutate(date=mdy(Date)) %>%
  dplyr::select(-Date)
```

Then we combined two tables into #comb# to help us do analyses. The #new_comb_ma# contains only data in MA.

```{r}
comb<-right_join(vaccine,case_death,by=c("date","state"))
#write.csv(comb,"bigtable.csv",row.names = FALSE)
new_comb_ma<-comb%>%
  filter(state=="MA")%>%
  arrange(date)
  #mutate(new_case=tot_case-lag(tot_case)) %>%
  #mutate(new_death=tot_death-lag(tot_death))%>%
  #mutate(new_adm=Administered-lag(Administered)) %>%
  #mutate(new_jj=Administered_Janssen-lag(Administered_Janssen)) %>%
  #mutate(new_moderna=Administered_Moderna-lag(Administered_Moderna))%>%
  #mutate(new_pfizer=Administered_Pfizer-lag(Administered_Pfizer))
```

# Exploratory Analysis

As for the EDA, we first focused on the data in MA only. We want to see the trend of ne

```{r}
# The number of new cases against date in MA
new_comb_ma%>%ggplot(aes(x=date))+
  geom_point(aes(y=new_case))+
  geom_vline(xintercept=as.Date("2021-07-03"),color="red")+
  geom_text(aes(x=as.Date("2021-04-30"),y=6700,label="Delta variant became dominant"))
```
The number of new death against date in MA
```{r}
new_comb_ma%>%ggplot(aes(date,new_death))+
  geom_point()+
  geom_vline(xintercept=as.Date("2021-07-03"),color="red")+
  geom_text(aes(x=as.Date("2021-08-30"),y=70,label="Delta variant became dominant"))
```

The change of new cases across all states
```{r}
comb%>%
  ggplot(aes(date,new_case,group=state))+
  geom_line(color='darkseagreen3',alpha=0.6)+
  geom_vline(xintercept=as.Date("2021-07-03"),color="red")+
  geom_text(aes(x=as.Date("2021-04-30"),y=40000,label="Delta variant became dominant"))
```
The change of new deaths across all states
```{r}
comb%>%
  ggplot(aes(date,new_death,group=state))+
  geom_line(color='darkseagreen3',alpha=0.6)+
  geom_vline(xintercept=as.Date("2021-07-03"),color="red")+
  geom_text(aes(x=as.Date("2021-08-30"),y=700,label="Delta variant became dominant"))
```

```{r}
# Distinguish states and represent new case using color
library(RColorBrewer)
comb %>% ggplot(aes(date, state,  fill = new_case)) + 
  geom_tile(color = "grey50") + 
  scale_x_continuous(expand = c(0,0)) + 
  scale_fill_gradientn(colors = brewer.pal(9, "Oranges"), trans = "sqrt") + 
  #geom_text(aes(x=as.Date("2021-08-30"), y=700, label="Delta variant became dominant")) + 
  theme_minimal() +  theme(panel.grid = element_blank()) + 
  xlab("") + ylab("")
```

Now, we are going to do some visualizations using maps.
```{r}
# Pull out us state map data frame
us_states <- map_data("state")

us_states %>% ggplot(aes(x = long, y = lat, fill = region, group = group)) + 
              geom_polygon(color = "white", fill = "grey") + 
              coord_fixed(1.3) +
              guides(fill = FALSE)
```

Before joining the map data with the CDC data, we need to make sure that the states match successfully with each other whenever possible.
```{r}
x <- case_death$state
case_death <- case_death %>% mutate(region = tolower(state.name[match(x, state.abb)]))
y <- vaccine$state
vaccine <- vaccine %>% mutate(region = tolower(state.name[match(y, state.abb)])) %>%
  mutate()
```

Then we can combine the two dataset using join functions.
```{r}
# Choose two dates (one before delta dominant, one after)
dates <- c(ymd("2021-06-01", "2021-09-01"))

# Filter case_death data for the two dates and join with us_states data frame.
comb_cd <- case_death %>% filter(date %in% dates)
comb_cd <- left_join(comb_cd, us_states, by = "region")

# Filter vaccine data for the two dates and join with us_states data frame.
comb_vac <- vaccine %>% filter(date %in% dates)
comb_vac <- left_join(comb_vac, us_states, by = "region")
```

First we explore the map of new cases and deaths.
```{r}
# Heatmap of new cases on June 1, 2021 and Sept. 1, 2021.
comb_cd %>% ggplot(aes(x = long, y = lat, group = group, fill = new_case)) + 
  geom_polygon(color = "white") + 
  coord_fixed(1.3) +
  facet_grid(date ~ .) +
  ggtitle("Map of new cases") +
  scale_fill_viridis_c(name = "New cases", option = "inferno", direction = -1, trans = "sqrt")
```

```{r}
# Heatmap of new deaths on June 1, 2021 and Sept. 1, 2021.
comb_cd %>% ggplot(aes(x = long, y = lat, group = group, fill = new_death)) + 
  geom_polygon(color = "white") + 
  coord_fixed(1.3) +
  facet_grid(date ~ .) +
  ggtitle("Map of new deaths") +
  scale_fill_viridis_c(name = "New deaths", option = "rocket", direction = -1, trans = "sqrt")
```

Then we explore the map of vaccination.
```{r}
# Filter case_death data on Sept. 1, 2021 and join with us_states data frame.
comb_cd <- case_death %>% filter(date == "2021-09-01")
comb_cd <- left_join(comb_cd, us_states, by = "region")

# Filter vaccine data for the two dates and join with us_states data frame.
comb_vac <- vaccine %>% filter(date == "2021-09-01")
comb_vac <- left_join(comb_vac, us_states, by = "region")
```

```{r}
# Heatmap of vaccine (regardless of brands) on June 1, 2021 and Sept. 1, 2021.
comb_vac %>% ggplot(aes(x = long, y = lat, group = group, fill = Administered)) + 
  geom_polygon(color = "white") + 
  coord_fixed(1.3) +
  ggtitle("Map of total vaccines for each state") +
  scale_fill_viridis_c(name = "Vaccination", option = "mako", direction = -1, trans = "log")
```

See the difference between vaccine brands.
```{r}
# Heatmap of vaccine (Moderna) on Sept. 1, 2021.
p1 <- comb_vac %>% ggplot(aes(x = long, y = lat, group = group, fill = Administered_Moderna)) + 
  geom_polygon(color = "white") + 
  coord_fixed(1.3) +
  ggtitle("Map of Moderna vaccinations") +
  scale_fill_viridis_c(name = "Vaccination", option = "mako", direction = -1, trans = "log")
```

```{r}
# Heatmap of vaccine (Pfizer) on Sept. 1, 2021.
p2 <- comb_vac %>% ggplot(aes(x = long, y = lat, group = group, fill = Administered_Pfizer)) + 
  geom_polygon(color = "white") + 
  coord_fixed(1.3) +
  ggtitle("Map of Pfizer vaccinations") +
  scale_fill_viridis_c(name = "Vaccination", option = "mako", direction = -1, trans = "log")
```

```{r}
grid.arrange(p1, p2, ncol = 1)
```



Classify the number of new cases into three categories (1: new case<=10000 on that day, 2: 10000<new case<=20000 on that day, 3: new case>20000 on that day)
Also classify the number of new deaths into three categories (1: new death<=200 on that day, 2: 200<new death<=400 on that day, 3: new death>400 on that day)
```{r}
clascomb<-comb%>%
  mutate(new_case_clas=ifelse(new_case<=10000,1,ifelse(new_case>20000,3,2)))%>%
  mutate(new_death_clas=ifelse(new_death<=200,1,ifelse(new_death>400,3,2)))
```

Randomly divide the table into one train set (70% data) and one test set(30% data)
```{r}
inTrain   <- createDataPartition(y = clascomb$new_case, p = 0.7, times = 1, list = FALSE)
train_set <- slice(clascomb, inTrain)
test_set  <- slice(clascomb, -inTrain)
```

Classification using decision tree model
```{r}
rpart_fit <- rpart(new_case_clas ~ Administered, data = train_set)
rpart_preds <- predict(rpart_fit, test_set)
rpart_res<-ifelse(rpart_preds<1.5,1,ifelse(rpart_preds>=2.5,3,2))
#rpart_preds
confusionMatrix(data = as.factor(rpart_res), reference = as.factor(test_set$new_case_clas),positive="1")
test_set%>%
  ggplot()+geom_point(aes(date,rpart_preds),color="red")+geom_point(aes(date,new_case_clas))
```
Classification using random forest model
```{r}
rf_fit <- randomForest(new_case_clas ~ Administered, data = train_set)
rf_preds <- predict(rf_fit, test_set)
rf_res<-ifelse(rf_preds<1.5,1,ifelse(rf_preds>=2.5,3,2))
#rpart_preds
confusionMatrix(data = as.factor(rf_res), reference = as.factor(test_set$new_case_clas),positive="1")
test_set%>%
  ggplot()+geom_point(aes(date,rf_preds),color="red")+geom_point(aes(date,new_case_clas))
```
Regression using random forest
```{r}
newcase_rf<-randomForest(new_case~Administered, data=train_set)
newcase_rf_pred<-predict(newcase_rf,test_set)
test_set%>%
  ggplot()+geom_point(aes(date,newcase_rf_pred),color="red")+geom_point(aes(date,new_case))
```


Divide the table into trainset and testset by date (cutoff point: 2021-07-31)
```{r}
trainset<-clascomb%>%
  filter(date <= as.Date("2021-7-31"))
testset<-clascomb%>%
  filter(date >as.Date("2021-07-31"))
#trainset
```

Machine learning using linear regression
```{r}
linearfit <- lm(new_case ~ Administered, data = trainset)
linear_pred <- predict(linearfit, testset)
data.frame(
  RMSE = RMSE(linear_pred, testset$new_case),
  R2 = R2(linear_pred, testset$new_case)
)
#testset%>%
#  ggplot()+geom_line(aes(date,rpart_preds1),color="red")+geom_point(aes(date,new_case))
ggplot(testset, aes(date, new_case) ) +
  geom_point() +
  stat_smooth(method = lm, formula = y ~ x)
```


Machine learning using random forest
```{r}
regr_rf<-randomForest(new_case~Administered, data=trainset)
regr_rf_pred<-predict(newcase_rf,testset)
testset%>%
  ggplot()+geom_point(aes(date,regr_rf_pred),color="red")+geom_point(aes(date,new_case))
plot(regr_rf)
sqrt(sum((regr_rf_pred - testset$new_case)^2) / nrow(testset))
```

